---
title: "Hot Spot Analysis"
subtitle: "Gi and Gi* foundations"
execute: 
  eval: false
  echo: true
format:
  revealjs:
    incremental: true   
---

```{r, eval = TRUE}
options(width = 78)
```


# Objective: 

identify clusters of high or low values in space

## LISA is going to help us

- local indicators of spatial association (LISA)
- decompose global statistics to the local level
- The Local Moran's I is the most famous

# Local Indicators of Spatial Association—LISA
Luc Anselin, 1995



## Which LISA for hot spots?

- Getis-Ord Gi and Gi*
- Local versions of the Getis-Ord Global G

:::{.aside}
_note: the $i$ subscript is a good hint that a measure is local_ 
:::

## Getis-Ord G

- G informs of the presence of high-high or low-low clustering

. . .

$$
G = \frac{\sum_{i = 1}^{n} \sum_{j = 1}^{n} w_{ij}x_ix_j}{\sum_{i = 1}^{n}\sum_{j = 1}^{n} x_i x_j}
$$

## Local Getis-Ord G ($G_i$)

- Most global statistics are a double sum     
  - first over $j$ and then $i$
- $G = \frac{\sum_{i = 1}^{n} \sum_{j = 1}^{n} w_{ij}x_ix_j}{\sum_{i = 1}^{n}\sum_{j = 1}^{n} x_i x_j}$
- Decompose the G into a LISA 
- $Gi = \frac{\sum_j w_{ij} x_j}{\sum_j x_j}$



## $G_i$

- The numerator is the spatial lag:  $\sum_j w_{ij} x_j$
- Denominator is the sum of all x values except $x_i$:  $\sum_j x_j$
- $G_i = \frac{\sum_j w_{ij} x_j}{\sum_j x_j}$
  
. . . 

> "ratio of the weighted average of the values in the neighboring locations to the sum of all values, not including the value at the location" ([GeodaCenter](https://geodacenter.github.io/workbook/6b_local_adv/lab6b.html#getis-ord-statistics))

. . . 

:::{.aside}
note: $G_i$ is often returned as a Z-score
:::

## Example

```{r, echo = FALSE, eval=TRUE}
library(sf)
library(sfdep)
library(tidyverse)

grid <- st_make_grid(cellsize = c(1, 1), n = 7, offset = c(0, 0)) |> 
  as_tibble() |> 
  st_as_sf() |> 
  mutate(
    id = row_number(),
    nb = st_contiguity(geometry),
    wt = st_weights(nb)
  )

nb <- grid[["nb"]]
wt <- grid[["wt"]]

set.seed(1)
y_sim <-  geostan::sim_sar(w = wt_as_matrix(nb, wt), rho = 0.78)

grd <- grid |> 
  mutate(x = y_sim - min(y_sim))

grd |> 
  ggplot(aes(fill = x)) +
  geom_sf(color = "black", lwd = 0.15) +
  scale_fill_viridis_c(option = "magma") + 
  theme_void()
```

## Example 
 
```{r, eval = TRUE, echo = FALSE}
grd |> 
  ggplot(aes(fill = x, label = round(x, 2))) +
  geom_sf(color = "black", lwd = 0.15) +
  geom_sf_text(color = "white") + 
  scale_fill_viridis_c(option = "magma") + 
  theme_void()
```

## Example data


```{r, eval = TRUE, echo = FALSE}
grd |> 
  select(x)
```
## The numerator

```{r, echo  = TRUE, eval = FALSE}
grd_nb <- grd |> 
  mutate(
    nb = st_contiguity(geometry),
    wt = st_weights(nb),
    x_lag = st_lag(x, nb, wt)
  )
```


## The numerator

```{r, echo = FALSE, eval = TRUE}
grd_nb <- grd |> 
  mutate(
    nb = st_contiguity(geometry),
    wt = st_weights(nb),
    x_lag = st_lag(x, nb, wt)
  )

grd_nb
```

## The numerator 

```{r, echo = FALSE, eval = TRUE}
ggplot(grd_nb, aes(fill = x_lag, label = round(x_lag, 2))) +
  geom_sf(color = "black", lwd = 0.15) +
  geom_sf_text(color = "white") + 
  scale_fill_viridis_c(limits = range(grd$x), option = "magma") +
  theme_void()
```

## The denominator 

- the of sum of all x values
- OR, spatial lag of a complete graph w/ binary weights
  - `st_weights(nb, style = "B")`
- Complete graph: all locations are connected
  - `st_complete_nb(n, diag = FALSE)`


## The denominator

. . .

```{r, eval = TRUE}
complete_nb <- st_complete_nb(n = nrow(grd), diag = FALSE)
complete_nb[[1]]
```

. . .

```{r, eval = TRUE}
denom <- st_lag(
  grd_nb$x, 
  complete_nb, 
  st_weights(complete_nb, style = "B")
)

denom
```



## The denominator 

```{r, eval = FALSE, echo = TRUE}
grd_nb |> 
  mutate(
    complete_nb = st_complete_nb(n()),
    wt_b = st_weights(complete_nb, style = "B"),
    sum_xj = st_lag(x, complete_nb, wt_b)
  ) |> 
  select(x, x_lag, sum_xj)
```

. . . 

```{r, eval = TRUE, echo = FALSE}
grd_nb <- grd_nb |> 
  mutate(
    complete_nb = st_complete_nb(n()),
    wt_b = st_weights(complete_nb, style = "B"),
    sum_xj = st_lag(x, complete_nb, wt_b)
  )  |> 
  select(x, x_lag, sum_xj)

grd_nb
```

## $G_i$

```{r, eval = TRUE}
grd_nb |> 
  mutate(
    complete_nb = st_complete_nb(n()),
    wt_b = st_weights(complete_nb, style = "B"),
    sum_xj = st_lag(x, complete_nb, wt_b),
    gi = scale(x_lag / sum_xj)
  ) |> 
  select(x, x_lag, sum_xj, gi)
```


## $G_i$ vs $G_i^*$

- The only difference is in neighbor matrix
- $G_i^*$ includes the observation $i$ in the calculation
- $G_i$ does **not**
- often rewritten as $G_i = \frac{\sum_{j\ne i} w_{ij} x_j}{\sum_{j \ne i} x_j}$
- I don't like this because they're the same except the `nb` list

## $G_i^*$

- _includes_ the observed value (focal observation)
- use `include_self(nb)`

. . . 


```{r, eval = TRUE, echo = FALSE}
gridExtra::grid.arrange(
  lattice::levelplot(
  nb_as_matrix(grd$nb),
  xlab = "",
  ylab = "",
  main = "self not included",
  scales = list(x = list(at = NULL), y = list(at = NULL))
  ),
  lattice::levelplot(
  nb_as_matrix(include_self(grd$nb)),
  xlab = "",
  ylab = "",
  main = "self included",
  scales = list(x = list(at = NULL), y = list(at = NULL))
  ),
  ncol = 2
)
```

## $G_i^*$ 

```{r, eval = TRUE}
grd |> 
  transmute(
    nb = include_self(st_contiguity(geometry)), # include the self 
    wt = st_weights(nb), 
    x_lag = st_lag(x, nb, wt), 
    # now we can sum all of x 
    denom = sum(x),
    gi = scale(x_lag / denom)
  )
```


## $G_i$  / $G_i^*$ interpretation

- High $G_i$ > 0: location with high value neighborhood
- Low $G_i$ < 0: location with low value neighborhood


## Hypothesis testing

- $H_0$ - observed value is not spatially dependent
- $H_1$ - observed value is spatially dependent
- $H_0$ - we would expect to find this value under **complete spatial randomness** (CSR)
- $H_1$ - we would __not__ expect to find this value under CSR




## resources:

- https://geodacenter.github.io/workbook/6a_local_auto/lab6a.html
